\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-05 paper.tex
Cut some more stuff

How nodes are interpreted, exactly, is application dependent.
Following \citet{hudson1983properties}, we can view nodes
as representing gametes, or we can imagine them representing,
for example, the genomes present in cells immediately before or after
some instantaneous event (Appendix~\ref{sec-cell-lineages-and-args}).
A node can represent any genome along a chain of cell divisions
or can be interpreted as representing one of the genomes of a
potentially long-lived individual. All these interpretations
are potentially useful, and equally valid under the assumptions
of the gARG encoding.
In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$,
and how we assign precise times will vary by application.
The topological ordering defined by the directed graph structure
and an arrow of time (telling us which direction is pastwards)
is sufficient for many applications, however,
and we assume node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined here is
in many ways simply a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% % This is definitely needed for the Nicks, who are always thinking
% % about the deeper processes and are quite happy being loosey-goosey
% % with representation details
In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process. The choice between using a gARG or an eARG to represent the general process of ancestral inheritance has 
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
important
consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This eARG encoding is the basis of the output formats created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph. These
represent the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While this approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it is inherently limited when
implemented literally. The eARG encoding explicitly models only
two different types of event; thus anything that is not a single crossover
recombination or common ancestor event must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. The gARG encoding described in the previous
section offers a simpler and more direct solution.

Aside from these practical challenges, there is also a deeper
issue with the implicit strategy of basing an ARG data structure on
recording events and their properties (e.g.\ the crossover breakpoint
for a recombination event). This approach assumes all events are 
\emph{knowable}, and does not
provide an obvious mechanism for either aggregating multiple events
or expressing uncertainty about them. While this is not a
problem when describing the results of simulations (where all details
are perfectly known), it is an issue when we wish to
formally describe the output of inference methods, particularly
as datasets approach the population scale.
As discussed in the introduction,
the precise details of all events in these vast ARGs can never be
known, so it is disadvantageous to use a data structure that enforces complete
precision.

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
% The gARG formulation avoids the need for such considerations,
% so is more naturally compatible with forming an ecosystem of interoperable
% inference and analysis methods.

@@

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient},
and the key distinguishing feature between the
``Big'' and ``Little'' ARG stochastic processes
(see Appendix~\ref{sec-big-and-little-arg}).
The ability to \emph{store} resolved ancestral material
is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain ancestral material without traversing the graph
pastwards from the sample nodes,
resolving the effects of recombination and common ancestor events.

>>>>

How nodes are interpreted, exactly, is application dependent.
Following \citet{hudson1983properties}, we can view nodes
as representing gametes, or we can imagine them representing,
for example, the genomes present in cells immediately before or after
some instantaneous event (Appendix~\ref{sec-cell-lineages-and-args}).
A node can represent any genome along a chain of cell divisions
or can be interpreted as representing one of the genomes of a
potentially long-lived individual.
In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$,
and how we assign precise times will vary by application.
The topological ordering defined by the directed graph structure
and an arrow of time (telling us which direction is pastwards)
is sufficient for many applications, however,
and we assume node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

% % % I feel like sophisticated readers would at this point be saying,
% % % ah, yeah, that's basically what an ARG is just in different words;
% % % what's the point of this paper?
% The genome ARG framework defined here is
% in many ways simply a clarification of existing treatments
% \cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
% adding concrete details to describe the
% differential inheritance of genetic material between genomes.
% % % This is definitely needed for the Nicks, who are always thinking
% % % about the deeper processes and are quite happy being loosey-goosey
% % % with representation details
% In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process. The choice between using a gARG or an eARG to represent the general process of ancestral inheritance has 
% % Again, need to remind the reader that there's an actual point
% % to all this getting stuck into the details.
% important
% consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This Event ARG (eARG) encoding is the basis of the output formats 
created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While this approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it has limitations.
The eARG encoding explicitly models only
two different types of event; thus anything that is not a single crossover
recombination or common ancestor event must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
% Graham asked for coalescences with GC papers, but this is the only one
% I know of?
conversion~\citep{wiuf2000coalescent} could 
be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. The gARG encoding described in the previous
section offers a simpler and more direct solution.

Aside from these practical challenges, there is also a deeper
issue with the implicit strategy of basing an ARG data structure on
recording events and their properties (e.g.\ the crossover breakpoint
for a recombination event). This approach assumes all events are 
\emph{knowable}, and does not
provide an obvious mechanism for either aggregating multiple events
or expressing uncertainty about them. While this is not a
problem when describing the results of simulations (where all details
are perfectly known), it is an issue when we wish to
formally describe the output of inference methods, particularly
as datasets approach the population scale.
% As discussed in the introduction,
% the precise details of all events in these vast ARGs can never be
% known, so it is disadvantageous to use a data structure that enforces complete
% precision.

% There is also a certain clarity gained by explicitly modelling nodes
% in the inheritance graph as genomes.
% Outside of the context of a
% % The point being, events are perfectly well defined in the models
% % but not in reality
% mathematical model, an ``event'' is a slippery concept.
% For example, \emph{which} genome along a chain of cell divisions should be
% regarded as the one where an event occurred,
% or whether multiple coalescences
% within a single individual should be regarded as one or multiple events are
% debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% % If we want to have an ARG software ecosystem then such fireside
% % discussions won't help.
% % The gARG formulation avoids the need for such considerations,
% % so is more naturally compatible with forming an ecosystem of interoperable
% % inference and analysis methods.

@@

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient}.
The ability to store resolved ancestral material
is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain ancestral material without traversing the graph
pastwards from the sample nodes,
resolving the effects of recombination and common ancestor events.
